\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-30 submission/manuscript.tex
edit for Jennas comment

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented {[}(12); . More notably, flawed ML practices are prevalent
such as using ML pipelines where there is no seperate held-out test
dataset to evaulate model performance, reporting few or only the best
outcomes of different randomizations of cross-validation and showing a
disregard for large differences between cross-validation and testing
performances as well as large confidence intervals of testing
performances (4, 13--19). Moreover, there is a lack of discussion on why
a particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11, 20--22). The lack
of transparency on modeling methodology and model selection negatively
impact model reproducibility and reliability. We need to strive toward
better machine learning practices by (1) implementing consistent and
reliable machine learning pipelines; (2) selecting ML models that
reflect the goal of the study as it will inform our expectations of
model accuracy, complexity, interpretibility and computational
efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite the simplicity, the
L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied only 0.021, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

For regularized logistic regression and SVM with linear kernel we used
L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero), removing features that could be
correlated with other important features, making it more difficult
include the relationships between the microbes and possibly dropping
features that are important in detecting SRNs. Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGboost, as well as an SVM with non-linear kernel.

We established a robust and reliable ML pipeline where we train and
validate each of the seven models {[}Figure 1{]}. We randomly split the
data into training/validation and test sets so that the
training/validation set consisted of 80\% of the full dataset while the
test set was composed of the remaining data {[}Figure 1{]}. Since the
cases are not uniformly represented in the data, the initial data-split
was stratified to maintain the overall label distribution on both the
training/validation and test sets. Training/validation set consisted of
393 patients (209 SRN), while the test set was composed of 97 patients
(52 SRN). The training/validation data was used for training purposes
and validation of hyperparameter selection, and the test set was used
for evaluation purposes. Validation of hyperparameter selection was
performed using 100 randomizations of five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over the 100 randomizations and selected the best performing
hyper-parameter setting to train the full training/validation dataset
{[}Figures S1 and S2{]}. We then used the held-out test set to evaluate
the prediction performance of each ML model. The data-split,
hyperparameter selection, training and testing steps were repeated 100
times to get a reliable and robust reading of model prediction
performance {[}Figure 1{]}.

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric. Random forest had significantly higher test AUROC
values than the other models for detecting SRNs. The median AUROC of
random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
SVM with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1 SVM with linear kernel and decision tree had
significantly lower AUROC values than the other ML models with median
AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
2{]}.

We evaluated the generalization performance of each ML model by
comparing their median cross-validation AUROC and median testing AUROC.
This difference should be low to suggest the model is not overfitting
despite the large number of features. The largest difference between the
two was 0.021 in L1 SVM with linear kernel followed by SVM with radial
basis function kernel and decision tree with a difference of 0.007 and
0.006, respectively {[}Figure 2{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost increase in
complexity as they decrase in interpretibility. We interpreted L1 and L2
SVM with linear kernel using the feature weights and L2 logistic
regression using regression coefficients of the trained models. We
examined the 5 predictive (highest weight) and protective (lowest
weight) features identified by the model. We calculated the mean weights
and coefficients of these features over the 100 data-splits. In the
three linear models, OTUs that had the largest mean weights and drove
the detection of SRNs belong to family \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629)
and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}. We explained the
feature importances in non-linear models using permutation importance on
the held-out test data. For the tree-based models, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most. Decision tree, random forest and
XGBoost models' predictive performance dropped from 0.6 base testing
AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}. The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
predictive performance.

@@

In this study we established a robust ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We showed the
importance of held-out test set and reporting the results of many
iterations of cross-validation and testing. Each data-split has a weight
that comes from random splitting. Seeing that testing and
cross-validation median differences being low show that the models are
generalizable and the pipeline works. Our results also show that we can
choose different models for different reasons. Getting feature
importances as weights and regression coefficients is the fastest and
easiest interpratation. Similarly, linear models require less
computational burden, fastest to train. THinking about the
applicability, and actionability of a model in clinical setting, we
might want faster results.

Suprisingly, l2 logistic regression does really well. computational
efficiency and interp of logistic vs random forest decision tree and l1
why so low?

@@

What happens with imbalanced data What happens with smaller datasets
Does the code work on any microbiome data? Why did we use only one
dataset? Why not validate on other data instead of held-out testing set?
Becuase we are comparing several methods on 1 dataset. What will the
future data look like?

@@

\paragraph{Data collection and study
population}\label{data-collection-and-study-population}

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (23). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic health was labeled by colonoscopy with adequate preparation and
tissue histopathology of all resected lesions. Patients with an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology were classified as advanced adenoma. Study had
172 patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(24--26).

\paragraph{Data definitions and
pre-processing}\label{data-definitions-and-pre-processing}

@@

\paragraph{Model training and
evaluation}\label{model-training-and-evaluation}

@@

>>>>

Machine learning (ML) modeling of the human microbiome is used to
identify the microbial biomarkers and aid in diagnosis of many chronic
diseases such as inflammatory bowel disease, diabetes and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes from bacterial abundances, but reliable ML models are
scarce in part to the flawed modeling methods. Furthermore, using black
box ML models has hindered the validation of microbial biomarkers. To
overcome these challenges, we benchmarked seven different ML models that
use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the accuracy,
interpretability, and computational efficiency of the following models:
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and extreme gradient boosting (XGBoost). The random forest
model was best at detecting CRC lesions with an AUROC of 0.695 but it
was slow to train (101.3 h) and hard to interpret. Despite its
simplicity, L2-regularized logistic regression followed random forest in
predictive performance with an AUROC of 0.680, and it trained the
fastest (12 min). This study showed that we should choose ML models
based on our expectations of accuracy, interpretability and
computational efficiency. It also established standards for modeling
pipelines of microbiome-associated ML models.

\subsection{Importance}\label{importance}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, flawed ML
practices are prevalent such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance, reporting
few or only the best outcomes of different randomizations of
cross-validation and showing a disregard for large differences between
cross-validation and testing performances as well as large confidence
intervals of testing performances (4, 13--19). Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11,
20--22). The lack of transparency on modeling methodology and model
selection negatively impact model reproducibility and reliability. We
need to strive toward better machine learning practices by (1)
implementing consistent and reliable machine learning pipelines and (2)
selecting ML models that reflect the goal of the study as it will inform
our expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization performance of each ML model. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite the simplicity, the L2-regularized
logistic regression followed random forest in performance. In terms of
computational efficiency, L2 logistic regression trained the fastest
(0.202 hours, std ± 0.019), while XGBoost took the longest (162.843
hours, std ± 3.986). We found that median cross-validation and testing
AUROC varied only 0.021, which highlights the importance of a seperate
held-out test set and consistent preprocessing of the data prior to
evaluation. Aside from evaluating generalization and classification
performances for each of these models, this study established standards
for modeling pipelines of microbiome-associated machine learning models.

@@

For regularized logistic regression and SVM with linear kernel we used
L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero), removing features that could be
correlated with other important features. Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGboost, as well as an SVM with non-linear kernel.

We established a robust and reliable ML pipeline where we train and
validate each of the seven models {[}Figure 1{]}. We randomly split the
data into training/validation and test sets so that the
training/validation set consisted of 80\% of the full dataset while the
test set was composed of the remaining data {[}Figure 1{]}. Since the
cases are not uniformly represented in the data, the initial data-split
was stratified to maintain the overall label distribution in both the
training/validation and test sets. Training/validation set consisted of
393 patients (209 SRN), while the test set was composed of 97 patients
(52 SRN). The training/validation data was used for training purposes
and validation of hyperparameter selection (i.e.~model selection), and
the test set was used for evaluation purposes. Validation of
hyperparameter selection was performed using repeated five-fold
cross-validation on the training/validation set {[}Figure 1{]}. Similar
to the initial data-split, five-fold cross-validation was also
stratified to maintain the overall label distribution on the training
and validation sets. We validated the cross-validation performances of
each hyperparameter setting over the 100 randomizations and selected the
best performing hyper-parameter setting in terms of AUROC to train the
full training/validation dataset {[}Figures S1 and S2{]}. We then used
the held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model prediction performance {[}Figure 1{]}.

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric {[}Figure 1{]}. Random forest had significantly
higher test AUROC values than the other models for detecting SRNs when
AUROC values were compared to the other six by Wilcoxon rank sum test (p
= 0.0011). The median AUROC of random forest was 0.695 (IQR 0.044). L2
logistic regression, XGBoost, L2 SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another. They had median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR
0.053), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1 SVM
with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

We compared the median cross-validation AUROC and median testing AUROC
for each model. This difference should be low to suggest the model is
not overfitting despite the large number of features. The largest
difference between the two was 0.021 in L1 SVM with linear kernel
followed by SVM with radial basis function kernel and decision tree with
a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.

We reported the testing AUROC values over 100 random splits of the full
dataset {[}Figure 1{]}. The testing AUROC values varied 0.23 on average,
depending on the data-split. For instance, the lowest AUROC value of the
random forest model was 0.59 whereas the highest was 0.81.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel using the feature weights and L2 logistic
regression using regression coefficients of the trained models. We
examined the 5 predictive (highest weight) and protective (lowest
weight) features identified by the model. We calculated the mean weights
and coefficients of these features over the 100 data-splits. In the
three linear models, OTUs that had the largest mean weights and drove
the detection of SRNs belong to family \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629)
and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}. We explained the
feature importances in non-linear models using permutation importance on
the held-out test data where we permuted groups of correlated features
(see methods). For the tree-based models, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most. Decision tree, random forest and
XGBoost models' predictive performance dropped from 0.6 base testing
AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}. The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
predictive performance.

@@

\begin{itemize}
\item
  In this study we established a robust ML pipeline to use 16S rRNA
  sequence counts to predict a binary health outcome.
\item
  We showed the importance of held-out test set and reporting the
  results of many iterations of cross-validation and testing. Each
  data-split has a weight that comes from random splitting. Seeing that
  testing and cross-validation median differences being low show that
  the models are generalizable and the pipeline works.
\item
  Our results also show that we can choose different models for
  different reasons. Getting feature importances as weights and
  regression coefficients is the fastest and easiest interpratation.
  Similarly, linear models require less computational burden, fastest to
  train. Thinking about the applicability, and actionability of a model
  in clinical setting, we might want faster results.
\end{itemize}

-This is a huge difference which shows that you can get ``lucky'' or
``unlucky'' depending on the random split. It is crucial to report all
the findings, many randomizations of these splits to get reliable
performance metrics.

\begin{itemize}
\item
  Suprisingly, l2 logistic regression does really well.
\item
  Computational efficiency and interp of logistic vs random forest
\item
  decision tree and l1 why so low?
\end{itemize}

@@

\begin{itemize}
\item
  What happens with imbalanced data
\item
  What happens with smaller datasets
\item
  Does the code work on any microbiome data?
\item
  Why did we use only one dataset?
\item
  Why not validate on other data instead of held-out testing set?
  Becuase we are comparing several methods on 1 dataset.
\item
  What will the future data look like?
\end{itemize}

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (23). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(24--26).

\textbf{Data definitions and pre-processing.}

@@

\textbf{Model training and evaluation.}

@@

\textbf{Permutation importance workflow.}
